---
title: "ATP Tennis results prediction - Project 1 - Descriptive Statistics"
authors: Francisco José Lozano Serrano, Antonio Miranda Escalada, Diego Suárez Villena
output:
  md_document:
    variant: markdown_github
---
```{r load_libraries, message=FALSE, echo =FALSE, warning = FALSE, results='hide'}
# Load libraries
packages_needed <- c("ggplot2", "GGally", "skimr", "dplyr", "fitdistrplus")
#install.packages(packages_needed)
lapply(packages_needed, require, character.only = TRUE)
```

```{r some prepro, echo=FALSE}
RawATP = read.csv("INPUT/atp_matches_with_stats_2016_17_no_null.csv")

# Drop ID, names, country of origin and date columns
drops <- c("tourney_id","p1_id", "p1_name", "p2_name", "tourney_name", "p2_id", "p1_ioc", "p2_ioc" ,"tourney_date", "match_num")
ATP <-RawATP[ , !(names(RawATP) %in% drops)]
numeric_variables <- unlist(lapply(ATP, is.numeric))
ATP$p1_win <- as.factor(ATP$p1_win)
ATP.Predictors <- ATP[ , !(names(ATP) %in% c("p1_win", "diff_points"))]
ATP.Predictors["diff_rank_points"] <- abs(ATP["p1_rank_points"] - ATP["p2_rank_points"])
ATP.Predictors["diff_age"] <- abs(ATP["p1_age"] - ATP["p2_age"])
ATP.Predictors["diff_rank"] <- abs(ATP["p1_rank"] - ATP["p2_rank"])

sapply(ATP.Predictors, class)
```

<br />

### FIRST DESCRIPTION

```{r skim, echo=TRUE, results='hide'}
# skim is not thought to be used on a markdown document. Run this on your R console
kable(skim(ATP))

```

First conclusions:

+ p1_hand: 28 unknown
+ p2_hand: 38 unknown
+ surface: 8 Nones

```{r tables_categorical_variables}
table(ATP$p1_hand)
table(ATP$p2_hand)
table(ATP$p1_win)
table(ATP$surface)
table(ATP$tourney_level)
```

<br />

+ tourney_level: London changes from A to F. Explore this variable. Only 13 observations with F tourney_level, which all correspond to the master's cup in London (there are also 27 observations of a tournament in London of category A)
```{r tourney_level_to_csv, echo = TRUE}
#write.csv(table(RawATP$tourney_name, RawATP$tourney_level),"tourney_levels.csv")
```

+ diff_points: small values? At least smaller than expected
+ draw_size: Should be related to the tournament level.
```{r tourney_level_draw_size}
table(ATP$drap1_size, ATP$tourney_level)
```

<br />

+ p1_age: ~normal distribution, no negative values, maximum in 39, minumun in 17 -> ok
+ p2_age: ~normal distribution, no negative values, maximum in 39, minumun in 17 -> ok
+ p1_rank: exponential distribution, no negative values
+ p2_rank: exponential distribution, no negative values
```{r histograms}
hist(ATP$p1_age, breaks = 20)
hist(ATP$p2_age, breaks = 20)
hist(ATP$p1_rank, breaks = 100)
hist(ATP$p2_rank, breaks = 100)
```
We may consider to log-transform the player rankings, to obtain a more Gaussian distribution.

<br />


+ p1_rank_points: exponential? gamma?log-transform.
+ p2_rank_points: exponential? gamma? log-transform.

```{r more histograms}
fit1_gamma = fitdist(ATP$p1_rank_points,"gamma",method="mme")
hist(ATP$p1_rank_points, freq = F, main="Histogram of p1_rank_points")
m = max(ATP$p1_rank_points)
grid=seq(0,m,1)
lines(grid,dgamma(grid,fit1_gamma$estimate[1],fit1_gamma$estimate[2]))
print(fit1_gamma$aic)

fit1_exp = fitdist(ATP$p1_rank_points,"exp",method="mme")
hist(ATP$p1_rank_points, freq = F, main="Histogram of p1_rank_points")
m = max(ATP$p1_rank_points)
grid=seq(0,m,1)
lines(grid,dexp(grid,fit1_exp$estimate[1],fit1_exp$estimate[2]))
print(fit1_exp$aic)

fit2_exp = fitdist(ATP$p2_rank_points,"exp",method="mme")
hist(ATP$p2_rank_points, freq = F, main="Histogram of p1_rank_points")
m = max(ATP$p2_rank_points)
grid=seq(0,m,1)
lines(grid,dexp(grid,fit2_exp$estimate[1],fit2_exp$estimate[2]))
print(fit2_exp$aic)

```

<br />

+ p1_ace_prob_20w: small values, more or less normal -> ok
+ p1_df_prob_20w: small values, more or less normal -> ok
+ p1_surface_win_prob_20w: -> seems ok
+ p1_svptWon_prob_20w: -> seems ok
+ p1_svptWon_prob_20w: -> seems ok

+ p2_ace_prob_20w: small values, more or less normal -> ok
+ p2_df_prob_20w: small values, more or less normal -> ok
+ p2_surface_win_prob_20w: -> seems ok
+ p2_svptWon_prob_20w: -> seems ok
+ p2_svptWon_prob_20w: -> seems ok

All these probabilities should show a difference between p1 and p2, because p1 has higher ranking than p2.
_ace and _svptWon could also be correlated, somehow. 

+ diff_rank: exponential -> log transform
+ diff_rank_points: exponential -> log transform
+ diff_age: gamma distribution. Decreasing but not exponentially

<br />

### MISSING VALUES

+ p1_hand: 28 unknown
+ p2_hand: 38 unknown
+ surface: 8 Nones

Input p1_hand and p2_hand unknowns to R level (most common class)
Input surface Nones to hard (most common class)
Input Carpet surface (only 3) to Grass (most similar in tennis terms)

```{r input_missing_values}
ATP[ATP$p1_hand == 'U',]$p1_hand <- 'R'
ATP[ATP$p2_hand == 'U',]$p2_hand <- 'R'
ATP[ATP$surface == 'None',]$surface <- 'Hard'
ATP[ATP$surface == 'Carpet',]$surface <- 'Grass'

ATP$p1_hand = droplevels(ATP$p1_hand, except = c('R', 'L'))
ATP$p2_hand = droplevels(ATP$p2_hand, except = c('R', 'L'))
ATP$surface = droplevels(ATP$surface)

table(ATP$p1_hand)
table(ATP$p2_hand)
table(ATP$surface)
```

### RELATIONSHIPS BETWEEN VARIABLES

<br />


#### Relationships between continuous variables 
```{r mega_plots_numerical, message=FALSE, warning=FALSE}

# Log scale exponential predictors
ATP.Predictors$p1_rank = log(ATP.Predictors$p1_rank)
ATP.Predictors$p2_rank = log(ATP.Predictors$p2_rank)
ATP.Predictors$p1_rank_points = log(ATP.Predictors$p1_rank_points)
ATP.Predictors$p2_rank_points = log(ATP.Predictors$p2_rank_points)
ATP.Predictors$diff_rank = log(ATP.Predictors$diff_rank)

# I have to add .5 to the diff_rank_points in order not to have -Inf in the log
ATP.Predictors$diff_rank_points = log(ATP.Predictors$diff_rank_points + 0.5)

# Add output variables to plot them with the numerical predictors
ATP.Predictors["diff_points"] <- ATP$diff_points
ATP.Predictors["p1_win"] <- ATP$p1_win

# I have to divide this mega-plot because all variables do not fit in one plot
nums.Predictors <- unlist(lapply(ATP.Predictors, is.numeric))

print(colnames(ATP.Predictors[,nums.Predictors]))

pn1 <- ggpairs(data=ATP.Predictors, columns = c(colnames(ATP.Predictors[,nums.Predictors])[2:11],
                                     "diff_points"),
               title="Numeric variables 2-11 + diff_points", 
               mapping=ggplot2::aes(colour = ATP.Predictors$p1_win, alpha = 0.7))
pn1

pn2 <- ggpairs(data=ATP.Predictors, columns = c(colnames(ATP.Predictors[,nums.Predictors])[2:6],
                                    colnames(ATP.Predictors[,nums.Predictors])[11:15]),
               title="Numeric variables 2-6,11-15", 
               mapping=ggplot2::aes(colour = ATP.Predictors$p1_win, alpha = 0.7))
pn2

pn3 <- ggpairs(data=ATP.Predictors, columns = c(colnames(ATP.Predictors[,nums.Predictors])[2:6],
                                     colnames(ATP.Predictors[,nums.Predictors])[16:19]),
               title="Numeric variables 2-6,16-19", 
               mapping=ggplot2::aes(colour = ATP.Predictors$p1_win, alpha = 0.7))
pn3

pn4 <- ggpairs(data=ATP.Predictors, columns = c(colnames(ATP.Predictors[,nums.Predictors])[7:15]),
               title="Numeric variables 7-15", 
               mapping=ggplot2::aes(colour = ATP.Predictors$p1_win, alpha = 0.7))
pn4

pn5 <- ggpairs(data=ATP.Predictors, columns = c(colnames(ATP.Predictors[,nums.Predictors])[7:11],
                                     colnames(ATP.Predictors[,nums.Predictors])[16:19]),
               title="Numeric variables 7-11,16-19",
               mapping=ggplot2::aes(colour = ATP.Predictors$p1_win, alpha = 0.7))
pn5

pn6 <- ggpairs(data=ATP.Predictors, columns = c(colnames(ATP.Predictors[,nums.Predictors])[11:19],
                                     "diff_points"),
               title="Numeric variables 11-19 + diff_points",
               mapping=ggplot2::aes(colour = ATP.Predictors$p1_win, alpha = 0.7))
pn6


# Remove output variables
ATP.Predictors["diff_points"] <- NULL
ATP.Predictors["p1_win"] <- NULL

```

<br />

###### Correlation
There are several variables highly correlated: 

+ p1_rank & p1_rank_points -> 0.98
+ p1_rank $ p2_rank_points -> 0.97
+ p1_surface_win_prob_20w & p1_win_prob_20w -> 0.82
+ p2_surface_win_prob_20w & p2_win_prob_20w -> 0.8
+ p1_rank & p1_win_prob_20w -> 0.74
+ p1_rank_points & p1_win_prob_20w -> 0.73

Others show moderate linear correlation: 

+ p2_rank & p2_win_prob_20w -> 0.65
+ p2_rank_points & p2_win_prob_20w -> 0.63
+ p1_svptWon_prob_20w & p1_ace_prob_20w -> 0.68
+ p2_svptWon_prob_20w & p2_ace_prob_20w -> 0.67

But most of the continuous variables are uncorrelated.

+ **WEIRD**! Strong correlation between diff_rank_points and p1_rank_points for p1_win == 0 (0.84), but not for p1_win == 1 (0.34).
More or less strong correlation between p2_win_prob_20w and p2_rank_points for p1_win == 0 (0.73), but not for p1_win == 1 (0.5).
More or less strong correlation between diff_rank_points and p2_rank_points for p1_win == 0 (0.53), but not for p1_win == 1 (0.16). CHECK COLORS!

<br />

###### Interesting remarks from scatter plots

+ The scatter plots of variable p2_svptWon_prob_20w (15) are always close to an horizontal line. This effect is less strong for p1_svptWon_prob_20w (11).

+ p2_rank_points (7) seems to divide more or less well the p1_win variable, with higher values of p2_rank_points (7) corresponding to p1_win == 1 and smaller ones to p1_win == 0.

<br />

###### Interesting remarks from distributions
+ p1_rank, p1_surface_win_prob_20w and p1_win_prob_20w behave slightly different for the two values of p1_win: the distribution of these variables do not completely overlap for the two values of p1_win.

<br />


#### Relationships between continuous variables and diff-points
Already in upper plots

<br />

#### Relationships between categorical variables and diff_points
Boxplots of diff_points by categories

```{r boxPlots_categorical_diff_points}
ggplot2::ggplot(ATP, aes(x=p1_hand, y=diff_points, color=p1_hand)) +
  geom_boxplot()
ggplot2::ggplot(ATP, aes(x=p2_hand, y=diff_points, color=p2_hand)) +
  geom_boxplot()
ggplot2::ggplot(ATP, aes(x=surface, y=diff_points, color=surface)) +
  geom_boxplot()
ggplot2::ggplot(ATP, aes(x=tourney_level, y=diff_points, color=tourney_level)) +
  geom_boxplot()
```

There is more difference in points in Grass mathces than in hard and Clay (in the mean), but the boxplot overlap is very high.

With respect to the tourney level, the means of the difference in points is higher for Grand Slam matches and Davis Cup matches, but in 2016 and 2017 (the years from where the data comes from) Davis Cup and Grand Slam matches were played up to 5 sets, so this is reasonable.

In Tourney_level:
+ A: All other mathces (smaller championships, etc.)
+ D: Davis Cups
+ F: finals (masters cup in London -> 13 matches)
+ G: Grand Slam
+ M: master 1000

<br />

#### Relationships between continuous variables and p1_win
Already in upper plots (color is p1_win)

<br />

#### Relationships between categorical variables and p1_win
Barplots of categorical variables by p1_win

```{r barplots_categorical_p1_win}
attach(ATP)
df_1 = data.frame(table(p1_win[p1_hand == "R"]), "R")
colnames(df_1) = c("p1_win", "Count", "p1_hand")
df_2 = data.frame(table(p1_win[p1_hand == "L"]), "L")
colnames(df_2) = c("p1_win", "Count", "p1_hand")
dftotal = rbind(df_2, df_1)
p1 <- ggplot(data=dftotal, aes(x=p1_win, y=Count, fill=p1_hand)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) 
p1

df_1 = data.frame(table(p1_win[p2_hand == "R"]), "R")
colnames(df_1) = c("p1_win", "Count", "p2_hand")
df_2 = data.frame(table(p1_win[p2_hand == "L"]), "L")
colnames(df_2) = c("p1_win", "Count", "p2_hand")
dftotal = rbind(df_2, df_1)
p2 <- ggplot(data=dftotal, aes(x=p1_win, y=Count, fill=p2_hand)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) 
p2

df_1 = data.frame(table(p1_win[surface == "Clay"]), "Clay")
colnames(df_1) = c("p1_win", "Count", "surface")
df_2 = data.frame(table(p1_win[surface == "Grass"]), "Grass")
colnames(df_2) = c("p1_win", "Count", "surface")
df_3 = data.frame(table(p1_win[surface == "Hard"]), "Hard")
colnames(df_3) = c("p1_win", "Count", "surface")
dftotal = rbind(df_2, df_1, df_3)
p3 <- ggplot(data=dftotal, aes(x=p1_win, y=Count, fill=surface)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) 
p3
table(surface[p1_win == 0])/dim(ATP[ATP$p1_win == 0,])[1]
table(surface[p1_win == 1])/dim(ATP[ATP$p1_win == 1,])[1]

```

There is not significant difference in the proportion of p1_wins depending on the hand or surface.

<br />

#### Relationships between categorical variables
Group bar plots
```{r bar_plots_categorical}
pgrey <- ggpairs(data=ATP, columns = c("p1_win", "p1_hand", "surface", "p2_hand", "tourney_level"), title="Categorical variables", mapping=ggplot2::aes(alpha = 0.9, position="dodge2"))
pgrey
```

<br />